Prefetching SPARQL query cacher

### 2016-05-20 dissertation/cache-proxy.tex
cost model details

>>>>

\subsection{Cost Model}\label{sec:costheuristics}

The final query execution is designed to be entirely cost-based,
i.e. many alternative plans are generated, but the decision on whether
to use a certain plan is entirely up to the cost model. While the
framework is flexible and could support further research into cost
models, I have considered this to be future work, and only a small
heuristic cost model is present in the current system.

However, since the combination of remotely executed SPARQL with a
local cache and Triple Pattern Fragments is rather complex, the cost
heuristics are fairly elaborate. Moreover, since the objective of the
study is to take load off of the remote endpoint, the heuristics is
intended to make the proxy work harder than the remote endpoint.

Throughout the system, the query planner takes the cost from the
models implementing the \pmodule{Attean::API::CostPlanner} role. A
key challenge is to balance the costs given by different models in the
system, and this is done in
\pmodule{AtteanX::Model::SPARQLCache::LDF}. Let us, however, start
from the bottom.

The cost of evaluating a Triple Pattern Fragment plan $C_{tpf}$ is estimated to
be 
\begin{equation}
C_{tpf} = 10 + \floor{990 \frac{n_{tpf}}{n_{tot}}} ,
\end{equation}
where $n_{tpf}$ is the estimated number of triples expected to be
matched by the triple pattern and $n_{tot}$ is the size of the
model. Thus, this cost will be an integer between 10 and 1000. The
reason we use a floor function is that the Attean cost model only
supports integers.

The cost of accessing the cache is assumed to be small compared to the
above, and is estimated as 
\begin{equation}
C_c = 2 + \floor{\log_{10}{n_c}} , 
\end{equation}
where $n_c$ is the number of triples in the cache that matches the
given triple pattern.

The cost of evaluating a remote SPARQL query is estimated to be the
number of subplans (usually quad patterns) incremented by one and
multiplied by 100 if there is a common variable between all the
subplans, or 1000 if there is not. This is to prevent the remote
endpoint from having to compute Cartesian joins as much as possible. A
remote SPARQL query with only one triple pattern is assumed to always
be more expensive than to evaluate than a Triple Pattern Fragment
(recall that we want to relieve the remote endpoint of stress), and so
is given the cost of 1001.

Next, we detail the cost of a join where one of more children in the
tree is a remote SPARQL plan. It is usually not wanted to execute
several remote queries, it is assumed that the number of HTTP requests
should be small, and that plans where the joins have been rotated and
coalesced by \pmodule{AtteanX::API::JoinRotatingPlanner} should be
less expensive. Therefore, such plans should be penalised, but by how
much depends on the type plans. Conventionally, for a nested loop join
plan, we multiply the cost of the subplans, increment the resulting
cost by one if the right hand plan is more expensive than the left
hand plan, and multiply by a factor of 10 if there are no shared
variables, to mildly penalise Cartesian joins. If the plan is a hash
join the difference is that, we add rather than multiply the cost of
the of the subplans, and multiply by a factor of 100 rather than 10 so
that if it is necessary to evaluate a Cartesian join, it is more
likely done using a nested loop join. Then, the resulting cost is
multiplied by the number of remote SPARQL Basic Graph Patterns in the
tree and the factor 1.2, and floored to the nearest integer.

Under the assumption that it is better to join remotely than to
possibly retrieve large amounts of data, a similar algorithm is used
to penalise join plans that contain many Triple Pattern Fragment
plans, by multiplying the cost of the join plan by the number of the
Triple Pattern Fragment plans in the tree.

Finally, if a plan has a Triple Pattern Fragment plan that has a
common variable with a remote SPARQL query, that plan is given an
added cost to 1000, to allow plans where the triple pattern has been
rotated into the BGP win, again to reduce the amount of data needed to
be transferred.
